---
title: "Replicating He et al. (2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replicating He et al. (2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette replicates Table 3 from He, Huang, Li, and Zhou (2023),
"Shrinking Factor Dimension: A Reduced-Rank Approach,"
_Management Science_, 69(9).

The table reports the total adjusted $R^2$ (%) of pricing 48 Fama-French
value-weighted industry portfolios using factors extracted by four
methods: the Fama-French factors directly (FF), PCA, PLS, and RRA.

## Setup

```{r}
library(sdim)

# Align dates: he2023_factors ends 12 months earlier than portfolio datasets
he2023_ff48 <- he2023_ff48vw[1:516, -1] / 100 - he2023_ff5$RF[127:642] / 100
G <- he2023_factors[1:516, -1] / 100

# First 6 columns are Fama-French 5 + momentum
f5 <- G[, 1:6]
```

## Replication

```{r}
nfact   <- c(1, 3, 5, 6, 10)
methods <- c("FF", "PCA", "PLS", "RRA")

total_r2 <- matrix(NA, nrow = length(methods), ncol = length(nfact))
rownames(total_r2) <- methods
colnames(total_r2) <- paste(nfact, "factors")

for (j in seq_along(nfact)) {

  k <- nfact[j]

  if (k <= 6) {

    total_r2["FF", j] <- eval_factors(he2023_ff48, f5[, 1:k])["TotalR2"]
    
  }

  fit_pca <- pca_est(target = he2023_ff48, X = G, nfac = k)
  total_r2["PCA", j] <- eval_factors(he2023_ff48, fit_pca$factors)["TotalR2"]

  fit_pls <- pls_est(target = he2023_ff48, X = G, nfac = k)
  total_r2["PLS", j] <- eval_factors(he2023_ff48, fit_pls$factors)["TotalR2"]

  fit_rra <- rra_est(target = he2023_ff48, X = G, nfac = k)
  total_r2["RRA", j] <- eval_factors(he2023_ff48, fit_rra$factors)["TotalR2"]

}
```

## Results

```{r}
round(total_r2, 2)
```

The RRA consistently achieves the highest total $R^2$ across all factor
counts, confirming the main finding of He et al. (2023): the
reduced-rank approach effectively shrinks factor dimension while
retaining pricing information.

## References

He, J., Huang, J., Li, F., and Zhou, G. (2023).
Shrinking Factor Dimension: A Reduced-Rank Approach.
_Management Science_, 69(9).
DOI: [10.1287/mnsc.2022.4563](https://doi.org/10.1287/mnsc.2022.4563)
