---
title: "Replicating Huang et al. (2022) — Table 4"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Replicating Huang et al. (2022) — Table 4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette replicates the IP growth results from Table 4 of Huang,
Jiang, Li, Tong, and Zhou (2022), "Scaled PCA: A New Approach to
Dimension Reduction," *Management Science*, 68(3).

The table reports out-of-sample R²_OS (%) of forecasting 1-month ahead
IP growth using PCA and sPCA factors extracted from 123 FRED-MD macro
variables. The benchmark is an AR model with SIC-selected lag order.

## Data

The package includes the authors' original data:

- `huang2022_macro`: 720 × 123 matrix of transformed FRED-MD predictors
  (January 1960 to December 2019).
- `huang2022_ip`: 720-vector of monthly IP growth (log-difference of IP
  index).

```{r}
library(sdim)

data(huang2022_macro)
data(huang2022_ip)

dim(huang2022_macro)
length(huang2022_ip)
```

## Methodology

The out-of-sample exercise uses an expanding window:

- **Initial estimation window**: January 1960 to December 1984 (300 months).
- **Out-of-sample period**: January 1985 to December 2019 (420 months).
- **Benchmark**: AR model with lag order selected by SIC (max lag = 1).
- **Factor models**: ARDL model with AR lags + 1 lag of PCA/sPCA factors.

For sPCA, the scaling regression uses the predictive relationship
y\_{t+1} ~ X\_{i,t}, and absolute slopes are winsorized at the 90th
percentile. The `spca_est()` function supports this directly: when
`length(target) < nrow(X)`, the first `length(target)` rows are used for
the scaling regression while all rows of `X` are standardized and used
for factor extraction.

## Out-of-sample loop

The full replication script is at `inst/replications/huang2022_table4.R`.
The OOS loop runs ~420 iterations and takes several minutes; below is a
condensed version:

```{r, eval = FALSE}
Z    <- huang2022_macro
y_ip <- huang2022_ip

M <- 300  # initial training window (Jan 1960 - Dec 1984)
N <- length(y_ip) - M
nfac_max <- 5

FC_AR <- FC_PCA <- FC_sPCA <- actual_y <- rep(NA, N)

for (n in seq_len(N)) {

  actual_y[n] <- y_ip[M + n]
  y_n  <- y_ip[1:(M + n - 1)]
  Z_n  <- Z[1:(M + n - 1), ]
  Zs_n <- scale(Z_n)  # standardize training window
  T_n  <- length(y_n)

  # AR benchmark (SIC selects from p = 0 or 1)
  # ... (see full script for details)

  # PCA factors
  pca_fit <- pca_est(X = Zs_n, nfac = nfac_max)
  z_pc    <- predict(pca_fit, Zs_n)

  # sPCA factors (predictive alignment + winsorization)
  spca_fit <- spca_est(
    target       = y_n[2:T_n],
    X            = Z_n,
    nfac         = nfac_max,
    winsorize    = TRUE,
    winsor_probs = c(0, 90)
  )
  z_spca <- predict(spca_fit, Z_n)

  # ARDL forecast using factors + AR lag
  # ... (see full script for details)
}
```

## Results

Running the full script produces:

```
R²_OS by number of factors (IP growth):
  K       PCA     sPCA
--------------------------
  1      8.97     9.65
  2      8.06    10.68
  3      8.22    11.09
  4      7.99    11.97
  5      7.88    13.17
--------------------------
```

With 5 factors, PCA achieves R²_OS = **7.88%** and sPCA achieves
R²_OS = **13.17%** — both matching the paper exactly. The sPCA
consistently outperforms PCA across all factor counts, confirming
that scaling predictors by their target-predictive slopes concentrates
forecasting-relevant information into the first few factors.

## Key `spca_est()` features used

1. **Predictive alignment**: passing a shorter `target` (T-1 elements)
   with the full `X` (T rows) ensures the scaling regression uses the
   predictive relationship y\_{t+1} ~ X\_{i,t} while factors are
   extracted from the full training window.

2. **Winsorization**: `winsorize = TRUE` with `winsor_probs = c(0, 90)`
   caps extreme scaling coefficients at the 90th percentile, matching
   the authors' MATLAB implementation.

3. **`predict()`**: projects the full training `X` onto the estimated
   sPCA loadings, correctly applying the training-window standardization
   and scaling.

## References

Huang, D., Jiang, F., Li, K., Tong, G., and Zhou, G. (2022). Scaled
PCA: A New Approach to Dimension Reduction. *Management Science*, 68(3),
1678--1695.
doi:[10.1287/mnsc.2021.4020](https://doi.org/10.1287/mnsc.2021.4020)
