---
title: "IPCA with the Grunfeld dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{IPCA with the Grunfeld dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the IPCA estimator using the Grunfeld (1958)
investment dataset, a classic panel of 11 US firms observed over 20 years
(1935--1954). Results are validated against the Python `ipca` package
(Kelly, Pruitt, and Su, 2019).

## The IPCA model

Instrumented PCA extracts latent factors from panel data where
asset-specific characteristics serve as instruments. The model is

$$r_{i,t} = \mathbf{z}_{i,t}^\top \boldsymbol{\Gamma} \mathbf{f}_t + \varepsilon_{i,t}$$

where $r_{i,t}$ is the return (here, investment) of asset $i$ at time $t$,
$\mathbf{z}_{i,t}$ is an $L$-vector of characteristics,
$\boldsymbol{\Gamma}$ is the $L \times K$ matrix of characteristic
loadings, and $\mathbf{f}_t$ is the $K$-vector of latent factors.

Estimation alternates between solving for $\mathbf{f}_t$ given
$\boldsymbol{\Gamma}$ and updating $\boldsymbol{\Gamma}$ given $\mathbf{f}_t$
(ALS), with an SVD normalization step to ensure
$\boldsymbol{\Gamma}^\top \boldsymbol{\Gamma} = \mathbf{I}_K$.

## Data preparation

The Grunfeld dataset ships with **sdim**. The dependent variable is gross
investment (`invest`) and the two characteristics are market value (`value`)
and capital stock (`capital`).

```{r}
library(sdim)

data(grunfeld)
str(grunfeld)
```

`ipca_est()` expects a $T \times N$ return matrix and a $T \times N \times L$
characteristics array:

```{r}
firms <- sort(unique(grunfeld$firm))
years <- sort(unique(grunfeld$year))
N  <- length(firms)
TT <- length(years)

ret <- matrix(NA_real_, TT, N, dimnames = list(years, firms))
Z   <- array(NA_real_, dim = c(TT, N, 2),
             dimnames = list(years, firms, c("value", "capital")))

for (i in seq_along(firms)) {
  
  idx <- grunfeld$firm == firms[i]
  ret[, i]  <- grunfeld$invest[idx]
  Z[, i, 1] <- grunfeld$value[idx]
  Z[, i, 2] <- grunfeld$capital[idx]
  
}

cat("ret:", nrow(ret), "x", ncol(ret), "\n")
cat("Z:  ", paste(dim(Z), collapse = " x "), "\n")
```

## Fitting IPCA

Extract one latent factor:

```{r}
fit <- ipca_est(ret, Z, nfac = 1)
print(fit)
summary(fit)
```

The returned object contains the characteristic loadings
($\boldsymbol{\Gamma}$, stored as `lambda`) and the estimated factors:

```{r}
# Gamma: how each characteristic maps to the factor
fit$lambda

# Factors over time
data.frame(year = years, factor = fit$factors[, 1])
```

## Validation against the Python `ipca` package

The Python `ipca` package implements the same ALS algorithm using the
Grunfeld dataset as its example. With `n_factors = 1` and no intercept,
the Python output is:

```{r}
py_gamma   <- c(0.99166014, 0.12888046)
py_factors <- c(
  0.1031968381, 0.0884489515, 0.0838496628, 0.0845069923, 0.0722523449,
  0.0995068155, 0.1228840058, 0.1422623752, 0.1197532025, 0.1179724004,
  0.1087561863, 0.1357521189, 0.1579348267, 0.1660545375, 0.1484923276,
  0.1586634303, 0.1596007400, 0.1759379247, 0.1921695585, 0.2111065868
)
```

Compare loadings and factors (sign-aligned):

```{r}
r_gamma <- as.numeric(fit$lambda)
r_factors <- as.numeric(fit$factors)

# Sign-align if needed
if (cor(r_gamma, py_gamma) < 0) {
  r_gamma   <- -r_gamma
  r_factors <- -r_factors
}

cat("Gamma max |diff|:  ", sprintf("%.2e", max(abs(r_gamma - py_gamma))), "\n")
cat("Factor max |diff|: ", sprintf("%.2e", max(abs(r_factors - py_factors))), "\n")
cat("Factor correlation:", sprintf("%.10f", cor(r_factors, py_factors)), "\n")
```

## Multiple factors

IPCA can also extract more than one factor. With $K = 2$ (both
characteristics contribute their own factor dimension):

```{r}
fit2 <- ipca_est(ret, Z, nfac = 2)
summary(fit2)
```

## References

Grunfeld, Y. (1958). The Determinants of Corporate Investment. Ph.D. thesis,
Department of Economics, University of Chicago.

Kelly, B. T., Pruitt, S., and Su, Y. (2019). Characteristics are Covariances:
A Unified Model of Risk and Return. *Journal of Financial Economics*, 134(3),
501--524. DOI: [10.1016/j.jfineco.2019.05.001](https://doi.org/10.1016/j.jfineco.2019.05.001)
