---
title: "Get started with sdim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with sdim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**sdim** implements five factor extraction methods for asset pricing and
macroeconomic forecasting:

| Function | Method | Reference |
|---|---|---|
| `pca_est()` | Principal Component Analysis (PCA) | He et al. (2023, MS) |
| `pls_est()` | Partial Least Squares (PLS) | He et al. (2023, MS) |
| `rra_est()` | Reduced-Rank Approach (RRA) | He et al. (2023, MS) |
| `spca_est()` | Scaled PCA (sPCA) | Huang et al. (2022, MS) |
| `ipca_est()` | Instrumented PCA (IPCA) | Kelly, Pruitt & Su (2019, JFE) |

All estimators return S3 objects with `print()`, `summary()`, and
`predict()` methods.

## Quick start

```{r}
library(sdim)

set.seed(42)
X   <- matrix(rnorm(200 * 20), 200, 20)
ret <- matrix(rnorm(200 * 30) / 100, 200, 30)
```

### PCA, PLS, and RRA

These methods take a multivariate target (T × N returns) and a matrix of
factor proxies (T × L):

```{r}
fit_pca <- pca_est(target = ret, X = X, nfac = 3)
fit_pls <- pls_est(target = ret, X = X, nfac = 3)
fit_rra <- rra_est(target = ret, X = X, nfac = 3)

print(fit_rra)
```

### Scaled PCA

sPCA takes a univariate target and scales each predictor by its OLS slope
on the target before extracting principal components. When
`length(target) < nrow(X)`, the first `length(target)` rows are used for
the scaling regression while all rows are used for factor extraction —
this supports the predictive alignment needed in out-of-sample
forecasting.

```{r}
y <- rnorm(200)

fit_spca <- spca_est(target = y, X = X, nfac = 3)
print(fit_spca)
```

### IPCA

IPCA extracts latent factors from panel data using time-varying
characteristics as instruments:

```{r}
n_periods <- 120; n_assets <- 50; n_chars <- 6
ret_panel <- matrix(rnorm(n_periods * n_assets) / 100, n_periods, n_assets)
Z <- array(rnorm(n_periods * n_assets * n_chars), dim = c(n_periods, n_assets, n_chars))

fit_ipca <- ipca_est(ret_panel, Z, nfac = 3)
print(fit_ipca)
```

## Prediction

Use `predict()` to project new data onto the estimated factor loadings:

```{r}
X_new <- matrix(rnorm(5 * 20), 5, 20)

# PCA projection
F_new <- predict(fit_pca, X_new)
dim(F_new)

# sPCA projection (standardizes newdata using training parameters)
F_spca_new <- predict(fit_spca, X_new)
dim(F_spca_new)
```

## Factor evaluation

Evaluate extracted factors using the metrics from He et al. (2023, §2.4):

```{r}
eval_factors(ret = ret, factors = fit_rra$factors)
```

## Bundled datasets

The package ships with datasets for replication:

- **`he2023_*`**: Seven datasets from He et al. (2023) — factor proxies
  and portfolio returns.
- **`huang2022_macro`**: 720 × 123 matrix of transformed FRED-MD
  predictors from Huang et al. (2022).
- **`huang2022_ip`**: IP growth target for the Huang et al. (2022)
  out-of-sample exercise.

See `vignette("he2023-table3")` and `vignette("huang2022-table4")` for
full replication examples.
